---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Assuming that the dataset is already unzipped in the working directory.Loading the data into current session and checking the summary.
```{r}
activity <- read.csv("activity.csv")
str(activity)
```
Coercing the class of the date variable from factor to date to facilitate future analysis.
```{r}
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?

day
Calculating the total number of steps taken per day and plotting it
```{r}
totalSteps <- tapply(activity$steps,activity$date,sum,na.rm=TRUE)
hist(totalSteps,col =  = "grey",main = "Total Steps taken per Day",xlab = "Total steps by date")
```
Following are the mean and median values in the same order
```{r}
mean(totalSteps,na.rm = TRUE)
median(totalSteps, na.rm = TRUE)
```


## What is the average daily activity pattern?

Time series plot of the 5 minute interval and the average number of steps taken, avaraged across all days
```{r}
avg_by_int <- tapply(activity$steps,activity$interval,mean,na.rm=TRUE)
plot(row.names(avg_by_int),avg_by_int,type="l",xlab="Time intervals (in minutes", ylab = "Average of Total Steps", main = "Time Series Plot of the average of Total Steps in a Day")
```
From the plot we see that the peak lies somewhere between 500-100 
```{r}
max_int <- max(avg_by_int)
x <- match(max_int,avg_by_int)
avg_by_int[x]
```
It can be seen that the 5-minute interval that contains the maximum number of steps is the time interval **835** with the corrosponding frequency *206.1698*

## Imputing missing values
Calculating the total number of missing values.
```{r}
sum(is.na(activity))
```
Replacing the Missing values by the mean of total steps taken per day. 
```{r}
avg <- avg_by_int
x <- which(is.na(activity))
new_activity <- activity
y <- new_activity$steps
y[x] <- avg
new_activity$steps <- y
```
Thus this new_activity dataset contains the imputed values as mean.
Plot of total steps taken per day with missing values imputed
```{r}
totalSteps_new <- tapply(new_activity$steps,new_activity$date,sum)
hist(totalSteps_new,fill = "grey",main = "Total Steps taken per Day",xlab = "Total steps by date")
```
Calculating Mean and Median with imputed data
```{r}
mean(totalSteps_new)
median(totalSteps_new)
```

Where previously with missing values removed the data was skewed to the left.With imputed data the graph more closely resembels a normal distribution .Imputed data gives a more accurate discription.
Intrestingly the mean and median of the imputed data are the same.


## Are there differences in activity patterns between weekdays and weekends?
Now to study the activity pattern across weekdays and weekends, We generate two factor variables "weekends" and "weekdays" to seggregate activities accordingly 
```{r}
library(dplyr)
library(plyr)
days <- weekdays(new_activity$date)
new_activity <- cbind(new_activity,days)
new_activity$days <- revalue(new_activity$days,c("Monday"="weekday","Tuesday"="weekday","Wednesday"="weekday","Thursday"="weekday","Friday"="weekday"))
new_activity$days <- revalue(new_activity$days,c("Saturday"="weekend","Sunday"="weekend"))
```
The new variable days is a factor variable with two levels - "weekdays" and "weekend" indicating whether a given date is a weekday or weekend.



```{r}
new_avg_by_int <- tapply(new_activity$steps,list(new_activity$interval,new_activity$days),mean)
                         library(reshape2)
new_avg_by_int <- melt(new_avg_by_int)
colnames(new_avg_by_int) <- c("interval","day","steps")
library(lattice)
xyplot(new_avg_by_int$steps ~ new_avg_by_int$interval | new_avg_by_int$day, layout=c(1,2),type="l",main="Time Series Plot of the Average of Total Steps (weekday vs. weekend)",xlab="Time intervals (in minutes)",ylab="Average of Total Steps")
